---
title: "Assignment 2: Coding Basics"
author: "Mazhar Bhuyan"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r, echo=TRUE}
#1.
Sequence_of_5 <- seq(1,55, by=5)
print (Sequence_of_5)
#2.First tried with separate codes for calculating mean and then print. But cat can do both at the same time.
mean(Sequence_of_5)
median(Sequence_of_5)
cat("Mean:", mean(Sequence_of_5), "\n")
cat("Median:", median(Sequence_of_5), "\n")
#3. Pretty similar to excel logic. Although I don't know why \n is necessary
if (mean(Sequence_of_5) > median(Sequence_of_5)) {
  cat("The mean is greater than the median.\n")
} else if (mean(Sequence_of_5) < median(Sequence_of_5)) {
  cat("The mean is less than the median.\n")
} else {
  cat("The mean is equal to the median.\n")
}
```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

```{r, echo=TRUE}
Student_Name <- c("A", "B", "C", "D") 

Test_Score <- c(40, 52, 62, 65)

Scholarship <- c(TRUE, TRUE, FALSE, TRUE)


#cat("Student Name:", A, B, C, D, "\n")
#cat("Test Scores:", test_scores, "\n")
#cat("On Scholarship:", on_scholarship, "\n")
# Didn't work out. 

cat("Student Names:", paste(Student_Name, collapse = ", "), "\n")
cat("Character Vector", "\n")
cat("Test Scores:", paste(Test_Score, collapse = ", "), "\n")
cat("Numeric Vector", "\n")
cat("Scholarship Status:", paste(Scholarship, collapse = ", "), "\n")
cat("Logic Vector", "\n")

```

6.  Label each vector with a comment on what type of vector it is.

```{r, echo=TRUE}
cat("Student_Name:", "Character Vector", "\n") # Character Vector
cat("Test_Score:", "Numeric Vector", "\n")   # Prints the scores
cat("Scholarship:", "Logic Vector", "\n") # Prints the scholarship status
```

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

```{r, echo=TRUE}
# Combining vectors into df
Student_Scholarship_Database <- data.frame(
  Student_Name = Student_Name, 
  Score = Test_Score, 
  Scholarship = Scholarship
)

print(Student_Scholarship_Database)
```

8.  Label the columns of your data frame with informative titles.

```{r, echo=TRUE}
colnames(Student_Scholarship_Database) <- c("Student Name", "Test Score", "Scholarship Status")


```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:A matrix contains only one type of data. In case of a data frame, like here, the data contains numeric, character logic vectors.

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail".

11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else`.

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r, echo=TRUE}
#10. Create a function using if...else

Evaluate_Score <- function(Score) {if (Score >50 ) {print("Pass")} else {print("Fail")}}

#11. Create a function using ifelse()

Evaluate_Score_Ifelse <- function(Score) {
  result <- ifelse(Score > 50, "Pass", "Fail")
  return(result)
}

#12a. Run the first function with the value 52.5
Evaluate_Score <- function(Score) {if (Score >52.5 ) {print("Pass")} else {print("Fail")}}


#12b. Run the second function with the value 52.5


Evaluate_Score_Ifelse <- function(Score) {
  result <- ifelse(Score > 52.5, "Pass", "Fail")
  return(result)
}

#13a. Run the first function with the vector of test scores
# I was hoping for an excel like on/off button. Need to consult

Student_Name <- c("A", "B", "C", "D") 

Test_Score <- c(40, 52, 62, 65)

Pass_Fail <-data.frame(Name = Student_Name, Score= Test_Score)

#Evaluate_Score <- function(Score) {if (Score >50 ) {print("Pass")} else {print("Fail")}}


#13b. Run the second function with the vector of test scores

Student_Name <- c("A", "B", "C", "D") 

Test_Score <- c(40, 52, 62, 65)

Pass_Fail <-data.frame(Name = Student_Name, Score= Test_Score)
#Evaluate_Score <- function(Score) {result <- ifelse(Score > 52.5, "Pass", "Fail") return(result)}

```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer: ifelse worked.

**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)
